---
title: "Box-Cox Transformation and Normalized Anomaly Rescaling <br> (trans-and-norman.R)"
output:
  html_document:
    theme: united
    css: SI-md-08.css
    highlight: haddock
    number_sections: yes
    fig_caption: yes
    toc: yes
    toc_float: false
    collapsed: no
---

```{r trans-and-zscore, echo=FALSE}
options(width = 105)
knitr::opts_chunk$set(dev='png', dpi=300, cache=TRUE)
pdf.options(useDingbats = TRUE)
```

# Introduction #

This script (`trans-and-norman.R`) loops over the individual site .csv files, and for each site determines the "optimal" (normalizing) Box-Cox transformation and then rescales the transformed data to normalized anomalies calculated over a specified base period.  There are two main parts to this script:  1) a set-up part that contains path and file names, along with the base-period specifiation (that change from run-to-run), and 2) the calculation part that generally does not change.

# Set up #

The first step is set various path names and base-period parameter values.  The `queryname` is used to compose file and pathnames, `datapath` specifies the folder where the input and output data reside, and `sitelistpath` and `sitelistfile` specify a particular list of sites to be processed.

```{r pathnames}
# trans-and-zscore.R

# 1-parameter Box-Cox transformation of charcoal quanities for a single site
# (alpha (shift parameter) is specified, lambda (power transformation parameter) is estimated)

# input .csv files should contain at least the variables "est_age" and "quant",
# 	identified by labels in a header row  

# paths for input and output .csv files -- modify as appropriate
queryname <- "v3i"
datapath <- "/Projects/GPWG/GPWGv3/GCDv3Data/v3i/"
sitelistpath <- "/Projects/GPWG/GPWGv3/GCDv3Data/v3i/v3i_sitelists/"
sitelist <- "v3i_nsa_globe"
```

The base period is specified by the beginning and end ages.  In this example, the beginning age is 250 (yr BP = 1700 CE) and the end age is 1950 (yr BP = 1 CE).  A label describing this base period (`basename`) is also defined.

```{r setbaseperiod}
# set base period ages (or read a baseperiod info file)
basebeg <- 250 
baseend <- 1950 
basename <- "nt2kb"
```

Set up a debugging (log) file.

```{r debug}
# debug/log file
debugpath <- "/Projects/GPWG/GPWGv3/GCDv3Data/v3i/v3i_debug/"
# if debug folder does not exist, create it
dir.create(file.path(debugpath), showWarnings=FALSE)
debugname <- "trans-and-norman_debug_nt2kb.txt"
# open the debug/log file
debugfile <- file(paste(debugpath, debugname, sep=""), "w")
```


# Calculation #

## Initial steps ##

First, various folders and filenames are created.  The `dir.create()` function is used to create the folders if they do not already exist.

```{r pathnames2}
# no changes below here
# various path and filenames
sitelistfile <- paste(sitelistpath, sitelist, ".csv", sep="")
sitecsvpath <- paste(datapath,queryname,"_sites_csv/",sep="")
transcsvpath <- paste(datapath,queryname,"_trans_csv/",sep="")
# if output folder does not exist, create it
dir.create(file.path(datapath, paste(queryname,"_trans_csv/",sep="")), showWarnings=FALSE)
statscsvpath <- paste(datapath,queryname,"_stats/",sep="")
# if output folder does not exist, create it
dir.create(file.path(datapath, paste(queryname,"_stats/",sep="")), showWarnings=FALSE)
statsfile <- paste(statscsvpath,queryname,"_",basename,"_stats.csv", sep="")
```

Read the list of sites to be processed.

```{r readsitelist}
# read list of sites
ptm <- proc.time()
sites <- read.csv(paste(sitelistfile), stringsAsFactors=FALSE)
nsites <- length(sites[,1])
print (nsites)
```

Define some vector arrays to store the statistics of the transformation (lambda, likelihood, etc.).

```{r statArrays}
# storage for statistics
sn_save <- rep(0,nsites); lam_save <- rep(0,nsites); lik_save <- rep(0,nsites)
tmean_save <- rep(0,nsites); tstdev_save <- rep(0,nsites)
```

## Main loop ##

Loop over the individual sites, doing the following:

1. compose the site .csv file name
2. read the input data
3. discard samples with mssing ages
4. discard samples with ages after 2020 CE
5. initial minimax rescaling (for historical reasons) (`minimax`)
6. set `alpha` the Box-Cox transformation shift parameter
7. maximum likelihood estimation of `lambda`
8. Box-Cox transformation of data
9. minimax rescaling `tall`
10. calculate mean over base period
11. calculate normalized anomalies (`normans`)
12. write out transformed data for this site


```{r mainLoop, eval=TRUE, results='hide'}
# main loop
for (j in seq(1,nsites)) {

  # 1. site .csv file name (input file)
  sitenum <- sites[j,1]
  sitename <- as.character(sites[j,5])
  siteidchar <- as.character(sitenum)
  if (sitenum >= 1) siteid <- paste("000", siteidchar, sep="")
  if (sitenum >= 10) siteid <- paste("00", siteidchar, sep="")
  if (sitenum >= 100) siteid <- paste("0", siteidchar, sep="")
  if (sitenum >= 1000) siteid <- paste(    siteidchar, sep="")
  inputfile <- paste(sitecsvpath, siteid, "_data.csv", sep="")
  print(j); print(sitenum); print(sitename); print(inputfile)

  # 2. read the input data
  sitedata <- read.csv(inputfile)
  nsamp <- length(sitedata[,1])
  nsampchar <- as.character(nsamp)
  writeLines(paste("Site",siteidchar,nsampchar,"samples", sep=" "), con = debugfile, sep = "\n")
  
  # 3. discard samples with missing (-9999) ages
  sitedata <- sitedata[sitedata$est_age != -9999,]
  
  # 4. discard samples with ages > -70
  sitedata <- sitedata[sitedata$est_age > -70,]

  # 5. initial minimax rescaling of data
  minimax <- (sitedata$influx-min(sitedata$influx))/(max(sitedata$influx)-min(sitedata$influx))
  
  # 6. set `alpha` the Box-Cox transformation shift parameter
  alpha <- 0.01  # Box-Cox shift parameter
  # alternative alpha: 0.5 times the smallest nonzero value of influx
  # alpha <- 0.5*min(sitedata$influx[sitedata$influx != 0])  

  # 7. maximum likelihood estimation of lambda
  # derived from the boxcox.R function in the Venables and Ripley MASS library included in R 2.6.1

  npts <- 201 # number of estimates of lambda
  y <- minimax+alpha
  n <- length(y)
  logy <- log(y)
  ydot <- exp(mean(logy))
  lasave <- matrix(1:npts)
  liksave <- matrix(1:npts)
  for (i in 1:npts) {
    la <- -2.0+(i-1)*(4/(npts-1))
    if (la != 0.0) yt <- (y^la-1)/la else yt <- logy*(1+(la*logy)/2*(1+(la*logy)/3*(1+(la*logy)/4)))
    zt <- yt/ydot^(la-1)
    loglik <- -n/2*log(sum((zt - mean(zt))^2 ))
    lasave[i] <- la
    liksave[i] <- loglik
    }

  # save the maximum likelihood value and the associated lambda
  maxlh <- liksave[which.max(liksave)]
  lafit <- lasave[which.max(liksave)]
  print (c(sitenum, maxlh, lafit))

  # 8. Box-Cox transformation of data
  if (lafit == 0.0) tall <- log(y) else tall <- (y^lafit - 1)/lafit

  # 9. minimax rescaling
  tall <- (tall - min(tall))/(max(tall)-min(tall))
  
  # 10. calculate mean of data over base period
  tmean <- mean(tall[sitedata$est_age >= basebeg & sitedata$est_age <= baseend])
  
  # 11. calculate "normans" normalized anomalies
  norman <- (tall-tmean)/tmean
  
  # 12. write out transformed data for this site
  siteout <- data.frame(cbind(sitedata[,1], sitedata$id_sample, sitedata$est_age,
    sitedata$depth, sitedata$influx, minimax, tall, norman))
  colnames(siteout) <- c("site_sample", "id_sample", "est_age", "depth", "influx", "influxmnx", "tall", "norman")

  outputfile <- paste(transcsvpath, siteid, "_trans_influx_", basename, ".csv", sep="")
  write.table(siteout, outputfile, col.names=TRUE, row.names=FALSE, sep=",")
  
  sn_save[j] <- sitenum
  lam_save[j] <- lafit
  lik_save[j] <- maxlh
  tmean_save[j] <- tmean
  
}
```

As the loop executes, one block of information for each site will be printed.

```{r printExample, echo=FALSE}
cat(" [1] 1 \n [1] 1 \n [1] \"Cygnet     \" \n [1] \"/Projects/GPWG/GPWGv3/data/v3plus/v3plus_sites_csv/0001_data.csv \" \n [1]    1.0000 -527.6608    0.2200 \n ...")
```

## Save transformation statistics ##

```{r writeStats}
# write out a file of statistics
stats <- data.frame(cbind(sn_save, lam_save, lik_save, tmean_save, tstdev_save))
colnames(stats) <- c("site", "lambda", "likelihood", "mean", "stdev")
write.table(stats, statsfile, col.names=TRUE, row.names=FALSE, sep=",")

proc.time() - ptm
```

